---
title: "eulerr: Area-Proportional Euler Diagrams with Ellipses"
subtitle: "Bachelor Thesis"
author:
  - name: "Johan Larsson"
  - name: "Peter Gustafsson"
affil:
  - name: "Lund University"
date: "`r Sys.Date()`"
automark: yes
colorlinks: yes
fontpack-libertine: yes
classoption:
  - headsepline=true
  - headings=standardclasses
biblatex: yes
biblio-style: ieee
header:
  - pos: ro
    next: eulerr
caption:
  - labelfont=bf
  - labelsep=period
  - font=small
setkomafont:
  - element: descriptionlabel
    commands: \normalfont\scshape\bfseries
bibliography: "eulerr.bib"
header-includes:
  - \usepackage[linesnumbered,lined,]{algorithm2e} 
output:
  komadown::scrartcl:
    keep_tex: yes
    fig_caption: yes
    pandoc_args: --biblatex
vignette: >
  %\VignetteIndexEntry{eulerr: Area-Proportional Euler Diagrams with Ellipses}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::knit_hooks$set(crop = knitr::hook_pdfcrop)

knitr::opts_chunk$set(
  echo = FALSE,
  warning = FALSE,
  message = FALSE,
  fig.width = 3,
  fig.height = 3,
  crop = FALSE
)

options(digits = 3)

library(eulerrPaper)
library(lattice)
library(latticeExtra)
library(gridExtra)
lattice.options(default.theme = list(fontsize = list(text = 8, points = 4)))

set.seed(1)

library(RConics)
library(eulerr)
```

# Background

Relationships between groups and sets are the focus of many scientific
disciplines. In biomedicine, for instance, the overlap between sets of genes
is often of interest. In epidemiology, the interactions between
diseases are frequently studied. Likewise, the social sciences are often
involved in studying demographics where the commonalities between various
groups are analyzed.

Visualizations of such relationships are central to understanding them.
The commonest visualization is Venn diagrams. Venn diagrams are a 
subset of Euler diagrams, which were originally proposed by Leonard Euler
(1707--1783) [@euler_1802]. Whereas Venn diagrams require that all
$2^n$ possible intersection are present -- even if they are empty -- Euler
diagrams stipulate no such requirement.

Euler diagrams can be area-proportional, which is the case for non-Venn
diagrams. Area-proportional diagrams are most easily understood as their
interpretations do not depend on any numbers. This paper will henceforth be
concerned only with such designs and the term *Euler diagram* will be
considered to refer to area-proportional diagrams.

Euler diagrams may be fashioned with any conceivable closed shape.  
Solutions have been developed for triangles [@swinton_2011],
rectangles [@swinton_2011], ellipses [@micallef_2014],
smooth curves, polygons [@swinton_2011], and
circles [@wilkinson_2012; @kestler_2008; @swinton_2011]. The latter is 
most common, and appears to be preferred for optimal
readability [citation]. Yet circles do not always lend themselves to
accurate representations. Consider, for instance the combination 
\[
\begin{gathered}
A = B = C = 2,\\
A \cap B = A \cap C = B \cap C = 1, \text{ and}\\
A \cap B \cap C = 0.
\end{gathered}
\]
There is no way to visualize this relationship perfectly with circles
yet with ellipses there *is* in fact a perfect solution
(Figure \@ref(fig:impossible)).

```{r impossible, fig.cap ="A set relationship depicted with circles and ellipses, showing that ellipses sometimes prevail where circles fail.", fig.width = 6}
fig_impossible()
```

With four intersecting sets, circular Euler diagrams are actually impossible,
given that $2^4=16$ regions are required but the circles can yield at most
14 unique intersections. This is not, however, the case with ellipses in that
they they may intersect in up to 4, rather than 2, points. All in all,
circles have three degrees of freedom: x-coordinate, y-coordinate, and
radius, whereas ellipses have five: x-coordinate, y-coordinate, semi-minor
radius, semi-major radius, and rotation.

Elliptical Euler diagrams have been successfully implemented in the 
**eulerAPE** software [@micallef_2014], yet only for three sets that moreover
need to intersect, the motivation for which being that Euler diagrams
with more than three sets often lack adequate solutions and that their
high complexity make implementations difficult [@micallef_2013].

As far as we know, analytical solutions to eulerr diagrams do not exist. All
existing implementations are instead based on varying numerical methods. 
Most of these use separate methods for the initial and final diagram
configurations. For instance,
@micallef_2013 uses uses a greedy algorithm, @wilkinson_2012 uses
multi-dimensional scaling with jacobian distances -- a constrained version of
which is used in @frederickson_2016 in combination with a greedy algorithm. The
latter two work only on pairwise relationships between the sets while the first
tries to optimize the three-way intersection between the three sets. All use
circles for the initial layout.

Diagrams with more than two sets require additional tuning, which is considered
in a final configuration step. This is always an optimization procedure with
a target loss function and an optimization procedure.

It is this difficulty that we have overcome with **eulerr**, which is the
first software to provide elliptical Euler diagrams for, theoretically, any
number of sets.

## Aim

The aim of this thesis is to present a method and implementation for
constructing and visualizing Euler diagrams for sets of any numbers with
ellipses.

# Method

Constructing a Euler diagram is in many ways analagous to fitting a
statistical model, in that you need

* data,
* a model to fit the data on,
* test to see that the model fits well, and
* a presentation of the result.

## Inputting data

Every Euler diagram begins with data that, in one way or another, is always
a description of set relationships. **eulerr** tries to be as friendly as
possible, catering to many forms of input:

Intersections and relative complements
:    $A \setminus B = 3 \quad B \setminus A = 2 \quad A \cap B=1$

Unions and identities
:    $A=4 \quad B=3 \quad A \cap B=1$

A matrix of logicals or binaries
:    $\begin{bmatrix}\bm{A} & \bm{B} & \bm{C} \\0 & 1 & 0 \\1 & 1 & 1 \\1 & 0 & 0 \\ \end{bmatrix}$

A list of sample spaces
:    $\begin{matrix} A = \{ab,\,bb,\,bc\}\\B = \{aa,\,bc,\,cc\}\\C = \{bb,\,bb,\,cc\} \end{matrix}$

A two- or three-way table[^1]
:    \begin{tabular}{lrr} \toprule Survived? & No  & Yes \\   \midrule Child & 52.00 & 57.00 \\   Adult & 1438.00 & 654.00 \\    \bottomrule\end{tabular}

[^1]: Data originally collected by the *British Board of Trade* in their
      investigation of the sinking of Titanic.

Whichever type of input is used, **eulerr** will eventually always convert it
to the first type, intersections and relative complements, since that is the
form used to compute the loss of the later optimizers but in the interim use
unions and identities, particularly when computing areas.

As an additional feature for the matrix form, the user may supply an additional
factor variable with which to split the data set before fitting a euler 
diagram to each split. This is offered as a convenience function for the user
since it may the case that the euler diagram solutions are relatively more
well-behaved after such a split.

## Fitting the euler diagram

The Euler diagram is fit in two steps: first, an initial configuration is
formed with circles using only the sets' pairwise relationships. Second, this
initial configuration is finalized, taking into account all relationships.

### Initial configuration

For our initial configuration, we work exclusively with circles. We begin by
mapping the disjoint set combinations to areas and, given these, figure out the
required pairwise distance between the sets to achieve a circle--circle overlap
matching these disjoint set combinations. We do this numerically, using the 
formula for a circle--circle overlap,

\begin{multline}
A = r_1^2\arccos\left(\frac{d^2 + r_1^2 - r_2^2}{2dr_1}\right) + 
r_2^2\arccos\left(\frac{d^2 + r_2^2 - r_1^2}{2dr_2}\right) - \\
\frac{1}{2}\sqrt{(-d + r_1 + r_2)(d + r_1 - r_2)(d - r_1 + r_2)(d + r_1 + r_2)},
\end{multline}

where $r_1$ and $r_2$ are the radii of the first and second circles
respectively and *d* the distance between them.

Although $r_1$ and $r_2$ are known, *d* is not, wherefore we approximate it
numerically. Here, our loss function is
the squared difference between *A* and the desired overlap, $(A-D)^2$,
which we then ptimize using the **R**'s built-in `optimize()`: a
"combination of golden section search and successive parabolic interpolation."
Convergence is fast and neglible in relation to our later, more demanding,
operations.

Given these optimal pairwise distances, we can proceed to the next step,
which is to position the circles representing the sets. This can be accomplished
in many ways; **eulerr** uses a method from the **venn.js** script
[@frederickson_2016], namely a constrained version of multi-dimensional scaling
(MDS), which in turn is an upgrade to a method from the **venneuler** package
[@wilkinson_2012]. **venneuler** tries to place disjoint and subset exactly
neck-in-neck and at the 
exact midpoint of the set respectively. Yet because we are indifferent about
where in the space outside (or respectively inside) the sets are placed, that
behavior becomes problematic since it might interfere with locations of other
sets that need to use that space.

The MDS algorithm from **venn.js** circumvents this by assigning a loss and gradient
of 0 when, for instance, the set relationsships *and* the candidate ellipses are
disjoint. Then, to optimize the pairwise relationsships between sets, **eulerr**
uses the following loss and gradient functions.

> \[\small
> \text{loss} = \sum_i \sum_j { {\begin{cases}
>     0 & \text{disjoint}(i, j)\\ 
>     0 & \text{subset}(i, j)\\ 
>     ((X_{i} - X_{j})^T(X_{i} - X_{j}) - D_{ij}^2) ^2  & \text{otherwise} \\ 
> \end{cases}}}
> \]
>
> \[\small
> \nabla f(X_{i}) = \sum_j {\begin{cases}
>      \vec{0} & \text{disjoint}(i, j)\\ 
>      \vec{0} & \text{subset}(i, j)\\ 
>      4 {((X_{i} - X_{j})^T(X_{i} - X_{j}) - D_{ij}^2)} (X_{i} -
>      X_{j}) & \text{otherwise} \\ 
> \end{cases}}\]
> *Source: [Better Venn Diagrams](http://www.benfrederickson.com/better-venn-diagrams/)
> by Ben Fredrickson, which includes a nice interactive demonstration.*

Fredrickson uses the *Polak--Ribière Conjugate Gradient Method* to optimize
the initial layout. In our experience this method occasionally runs into
local minima, which is why
we have opted to use `nlm()` from the **R** core package `stats`: a 
translation from FORTRAN code developed by @schnabel_1985, which uses a mix
of Newton and Quasi-Newton algorithms. `nlm()` uses the Hessian in addition
to the gradient, which we presently compute numerically due to a bug
in the software implementation wherein the analytic Hessian is updated
incorrectly.

This initial configuration will work perfectly for any 1--2 set combinations
and as well as possible with 3 sets (if we use circles). But for all other
combinations there is usually a need to fine tune the configuration.

### Final configuration

So far, we have only worked with pairwise relationships and areas. To 
test and improve our layout, however, we now need to account for all
relationships and consequently all the various intersections between the
shapes we use in the diagram -- initially, we restriced ourselves to circles but
we now extend our reach also to ellipses.

As it turns out, computing the overlaps of ellipses is not a trivial task.
In fact, most of the methods we studied rely on
approximations by, for instance, quad-tree
binning [@wilkinson_2012] or polygon intersections [@kestler_2008]. These
methods are both imperfect [@micallef_2013] and, given that the
computations may have to run for vast numbers of iterations, prohibitive
performance-wise.

Other methods [@micallef_2013; @frederickson_2016], however, use algorithms
that compute the overlap areas exactly. But unlike most approximative
types, all the exact methods require that we first find the points
of intersection between the ellipses. **eulerr**'s approach to this is based
on a method outlined by @richter-gebert_2011, which is based in 
*projective geometry* rather than euclidean. Then, to find the 
intersections, we

* convert the two ellipses from canonical form to matrix notation. The 
canonical form of a rotated ellipse is given by
\[
\frac{((x-h)\cos(\phi)+(y-k)\sin(\phi))^2}{a^2}+\frac{((x-h) \sin(A)-(y-k)
  \cos(\phi))^2}{b^2} = 1,
\]
where *phi* is the counter-clockwise angle from the positive x axis to the 
semi-major axis *a*. *b* is the semi-minor axis whilst *(h, k)* is the center
of the ellipse. This is then converted to the matrix form
\[
E = \begin{bmatrix}A & B/2 & D/2 \\
                   B/2 & C & E/2 \\
                   D/2 & E/2 & F
    \end{bmatrix},
\]
which may be used to represent any conic. We then
* form a degenerate conic from the solution to the system of these two conics,
* split this degenerate conic into a pencil of two lines, and finally
* intersect the remaining conic with this pencil, yielding 0 to 4 intersection
points (\@ref(fig:intersection)).

```{r intersection, fig.cap = "The process (from left to right) used to intersect two ellipses, here yielding 4 points.", fig.width = 6, fig.height = 2}
fig_intersection()
```

After we have discovered all of the intersection points between the ellipses,
we find the area of their overlap by examining the subset of intersection
points that were formed from the presently examined ellipses that are also
within all of these ellipses. These intersection points form a convex polygon
plus elliptical segments (Figure \@ref(fig:polyarea).

```{r polyarea, fig.cap="The overlap area between three ellipses is the sum of a convex polygon (in grey) and 2--3 ellipse segments (in blue).", fig.height = 2.4}
fig_polyarea()
```

Since the polygon section is always convex, it is trivial to find its area using
the the triangle method. To find the areas of the elliptical segments, we use
an algorithm developed from @eberly_area_2016, in which we
integrate the ellipse between the two points making up the start and end of
the elliptical segment. This gives us the elliptical sector from which we
subtract the triangle part to find the segment area
(Equation \@ref(eq:segmentArea)).
\begin{equation}
A(\theta_0, \theta_1) = F(\theta_1) - F(\theta_1) -
\frac{1}{2}|x_1y_0 - x_0y_1|,
(\#eq:segmentArea)
\end{equation}
\[
\text{where } F(\theta) = \frac{a}{b}\left[ \theta -
\arctan{\left(\frac{(b - a)\sin{2\theta}}{b + a +(b - a )\cos{2\theta}} \right)}
\right]
\]
To our knowledge, this method was first published for circles in a blog post 
[@frederickson_2013] and has been extended in **eulerr** to also work for
all ellipses.

With this in place, we are now able to compute the areas of all the set
intersections and their relative complements up to numerical precision, which we
take advantage of in the final step of the configuration. 

We now feed the initial layout computed in [Initial configuration] to the
`nlm()` optimizer. We proceed to let the ellipses rotate and relax the
criterion that the semi-major and semi-minor axes are the same (as in a circle).
This gives us five parameters to optimize per set. For each iteration of the
optimizer, the areas of the intersections are analyzed and a measure of 
loss computed. A a loss function, we use the sum of squared differences
between the set intersections plus their relative complements and
the respective areas of the diagram (Equation \@ref(eq:loss)).
\begin{equation}
\sum_{i=1}^{n} (A_i-\omega_i)^2
(\#eq:loss)
\end{equation}

## Goodness of fit

Before we get to plotting the solution, it is useful to know how well the
fit from **eulerr** matches the input. Sometimes euler diagrams are just not
feasible, particular for combinations with many sets, in which case we should
stop here and look for another design to visualize the set relationships.

It is not, however, obvious what it means for a euler diagram to "fit well". 
**venneuler** uses a metric called *stress*, which is defined as
\[
\frac{\sum_{i=1}^{n} (y_i - \hat{y}_i) ^ 2}{\sum_{i=1}^{n} y_i ^ 2}
\]
where $\hat{y}_i$ is an ordinary least squares estimate from the regression of
the fitted areas on the original areas that is being explored during
optimization.

Meanwhile, **eulerAPE** [@micallef_2014] uses *diagError*:
\[
\max_{i = 1, 2, \dots, n} \left| \frac{y_i}{\sum y_i} -
  \frac{\hat{y}_i}{\sum \hat{y}_i} \right|
\]

Both metrics are given the user after the diagram has been fit, together
with a table of residuals.

```{r, fig.cap = "A plot with circles."}
combo <- c("A" = 1, "B" = 1, "C" = 1,
           "A&B" = 0.5, "A&C" = 0.5, "C&B" = 0.5)

fit1 <- euler(combo)
fit1
plot(fit1, counts = TRUE)
```

It is clear that this is not a good fit, which we can find out just by
looking at the plot. This is a good example of when ellipses come in handy.

```{r, fig.cap = "A plot with ellipses."}
fit2 <- euler(combo, shape = "ellipse")
fit2
plot(fit2, counts = TRUE)
```

Much better.

[TODO]: # What is a high and low diag_error?

## Visualization

Euler diagrams are naught without visualization. Here, **eulerr**
interfaces the elegant Lattice graphics system [@sarkar_2008] to grant the
user extensive control over the output, and allow for facetted plots in
case such a design was used in fitting the euler configuration.

In this case, layout out the diagram becomes considerably more involved. Finding
a reasonable spot for the text inside the diagram only lends itself to an
easy solution if the shape of the intersection has a center-of-gravity inside
ellipse, in which case an average of some of the points might suffice. This
is often not the case, however, and we need a better solution. Specifically,
what we need is a method to find the point inside the circle overlap for the
counts and circle complement to the intersection for our labels.

So far, we have not been able to derive at an analyitcal solution for
finding a good point, or for that matter a reliable way of finding *any* point
that is in the required intersection or complement. As is often the case,
the next-best thing turns out to be a numerical one. First, we locate a 
point that is inside the required region by spreading points across one of the
discs involed in the set combination. To spread points uniformly, we use
*Vogel's method* [@arthur_2015; @vogel_1979]
\[
\left( p_k = (\rho_k, \theta_k) = \left( r \sqrt{\frac{k}{n}},\, \pi (3 - \sqrt{5})(k - 1) \right) \right)_{k=1}^n,
\]
which is actually based on the golden angle.

```{r, fig.cap = "Spreading points on a disc with Vogel's method."}
n <- 500
seqn <- seq(1, n, 1)
theta <- seqn*pi*(3 - sqrt(5))
rad <- sqrt(seqn/n)
x <- rad*cos(theta)
y <- rad*sin(theta)

xyplot(x ~ y, aspect = "iso", pch = 20, xlab = "", ylab = "",
       col = 1,
       par.settings = list(axis.line = list(col = "transparent")),
       scales = list(draw = FALSE))
```

After this, we scale, translate,
and rotate the points so that they fit the desired ellipse.

---
TODO: Visualize this step.
---

After we've spread our points throughout the ellipse and found one that
matches our desired combination of ellipses/sets, we then proceed to optimize
its position numerically. For this, we use version of the
*Nelder--Mead Method* [@nelder_1965] which we've translated from Matlab
code by @kelley_1999 and customized for **eulerr**
(in particular to make sure that
the simplex does not escape the intersection boundaries since we for this
problem *want* the local minimum).

## Coloring
Per default, the ellipses are filled with colors. The default option is to
use an adaptive scheme in which colors are chosen to provide a balance
between dinstinctiveness, beauty, and consideration for the color deficient.
The color palette has been generated from
[qualpalr](https://CRAN.R-project.org/package=qualpalr)
(developed by the author), which automatically generates qualitative color
palettes based on a model of color perception.

# Results

## Consistency

## Accuracy

## Performance

Being that we're optimizing over 5 parameters for every ellipse (or 3 in the case
of circles) -- nevertheless, we profit from our implementation
in the C++ programming language via the **R** packages **Rcpp**
[@eddelbuettel_2011] and **RcppArmadillo** [@eddelbuettel_2014].

# Discussion

# Layout
Since the optimization steps are unconstrained, we run the risk of
ending up with dispersed layouts. To fix this, we use the SKYLINE-BL 
rectangle packing algorithm [@jylanki_2010] to pack the disjoint clusters of
ellipses (in case there are any) into a heuristically chosen bin.

At the time of writing this algorithm is crudely implemented -- for 
instance, it does not attempt to rotate the rectangles (boundaries for the
ellipses) or attempt to use. Since we're dealing with a rather simple version
of the rectangle packing problem, however, it seems to do the trick.

## Labelling


# References
